Impact on End-users by ISP IPv6 Deployment

### 2023-01-03 paper.tex
checked with grammarly

    \begin{abstract}
        This paper presents an overview of different techniques to cope with the IPv4 address exhaustion. The impact on the end-users is on focus. First some early NAT based solutions are introduced. Later dualstack is discussed. At the end IPv6 only solutions, deployed by ISPs today, are presented. It turns out that the end-user, especially the technical experienced one, can experience benefits with IPv6. With IPv6 the internet gets back to its routs. All devices can talk to each other and central server are not mandatory anymore to establish communication.
    \end{abstract}

@@

    In the late 90s it became clear, that the IPv4 address space is not sufficient to meet future demand. The popularity of the Internet especially in the end-user marked was heavily underestimated at the creation of IPv4. There are approximately 4 billion IPv4 addresses. Today there are over 8 billion humans on earth. That means that two humans need to share a single IPv4 address, not considering servers. Today, a typical family has multiple devices per person, including TVs, gaming consoles, smart phones, personal computers and more. This led to the assignment of the last available IPv4 block in the early 2010s. With the rise of smart home technology and IoT devices, the demand for IP addresses per household will not stop to increase. In this paper I will present a nearly complete overview of the different ideas to slow down the inevitable IPv4 exhaustion. Expect for the last solution, none of these is a long term one or even tries to solve the problem itself. ISPs tried to keep cost as low as possible. This led to keeping IPv4 addresses as long as possible.\\\\First, ISPs started to assign only one public IP per customer. The customer had to use NAT44 in order to provide all his end devices with an internet connection. The end-user was the one getting a worse experience. Over time, assigning only one IPv4 per customer was too much for the ISPs. Especially with the rise of mobile internet and an even higher demand for addresses, IPv4 was not sufficient anymore. ISPs tried to provide more IP addresses, by assigning only private IPv4 addresses to customers, leading to NAT444. NAT itself has problems, which will be discussed in this paper. NAT444 makes the problems even worse.\\\\A few years later dualstack was provided by ISPs. Dualstack introduced the next generation of IP addresses, called IPv6 addresses. Compared to IPv4 addresses which are a 32 bit wide, IPv6 addresses are a 128 bit wide. IPv6 completely solves the problem of address exhaustion. Unfortunately IPv6 is not backwards compatible to IPv4. Thus all infrastructure needs to be renewed. The Internet is fully decentralized. It is not possible to perform such a transition on one particular day. Dualstack implements both, IPv4 and IPv6. Devices and servers supporting IPv6 can communicate over IPv6. In the case that one device is not yet capable of IPv6, IPv4 acts as a fallback. This is a perfect transition technology, but it is no solution for the IPv4 exhaustion problem. Each end-user still needs an public or private IPv4. The problem is not solved.\\\\Lately, ISPs started to provide IPv6 only. This is also common in mobile networks. The devices natively speak to servers over IPv6. For legacy IPv4 servers, there are different translation mechanisms. This paper will take a closer look into NAT64 combined with DNS64. Moreover I will present 464XLAT. Both are similar to a NAT44, but translating between IPv6 and IPv4. The interesting thing is that these translations happen at the ISP, so the end-user only uses with IPv6. This has advantages for the end-user. Though, some legacy applications that depend on IPv4 cannot be used. For example, Github is not reachable over Ipv6. The DNS does not return a \textit{AAAA} record.\\\\At the end of this paper, tunneling IPv6 in IPv4 is presented. This happens only at ISP level and indirectly influences the end-user too. Mainly in better path availability, thus seamless internet service. There are various technologies. 6in4 and 6rd will be introduced shortly.\\\\In the last few years, the adoption of IPv6 has increased rapidly. Above 40 percent of all google users access the site over Ipv6 \cite{GoogleIPv6}. This statistics really shows the trend of ISPs worldwide, deploying IPv6 only or dualstack. Nearly all mobile clients use IPv6 today and are able to interact with the Internet without notable disadvantages. Cisco predicts that the transition to IPv6 is finished in 2028 \cite{Hughes2022_C05}. This means that nearly all end-users and servers world-wide communicate over IPv6. That has noticable advantages for the end-user regarding security, because firewalls can be managed easier. In addition it opens up entirely new options for innovative peer to peer technologies. Hosting a Server on a smart phone might sound strange today, but could be a usual thing in 10 years from now. The end-user is the one benefiting the most from unique, world-wide routable IP addresses.\\\\Now I am going to turn back the time a bit and we are focusing on the problem, emerging in the late 90s. It becomes clear that the IPv4 address space is not big enough. ISPs are searching for a solution. The main goal is to assign the same public, globally routable IPv4 address to multiple devices. This is not meant for a long period of time, but much more as a quick fix, until the new standard IPv6 is developed.

@@

    This section presents three different solution to extend the lifetime of IPv4: NAT44, A+P and NAT444. All these solutions aim to share one single IPv4 address between multiple end devices. It is important to understand theses technologies in order to correctly estimate the impact of IPv6 deployment by ISPs. Evaluating modern solutions alone could lead to the impression, that these solutions are better or worse than they really are. Thus this paper starts with NAT44.

@@

    Everybody can use the IPv4 addresses in these blocks. Private IPv4 addresses will never be routed in the internet. Most end-users use the third option for their home network, but only with a /24 network. The third block remains 0 all the time.Thus they can use 254 different devices which is sufficient for most users. All these 254 devices share one public IPv4. All of them share the $2^{16} = 65536$ ports of the public IPv4 too. From the outside you cannot decide whether only one device or multiple devices are behind the NAT.

@@

    In the example Figure \ref{fig:nat_44} is a NAT-Translation table. As you can see, the PC with IP \textit{192.168.0.10} cannot use all ports, because the PC with the IP \textit{192.168.0.12} uses the Port \textit{7035}. All ports are shared. This can lead to problems as we discuss later. NAT44 was the first solution to the IPv4 exhaustion problem. It is an easy fix, because the ISP has nothing to change, except now only delivering one single IP per customer. The end-user needs a NAT compatible router.

@@

    \textit{A+P} means \textit{Address Plus Port}. Over the time NAT44 was not a sufficient solution anymore. The demand for IPv4 rose and ISP were not able to provide one single public, globally routable IPv4 per customer. The idea is still similar to NAT44, but with \textit{A+P} each costumer gets one IPv4 and a designated port range. With that system it is possible to share one IP to theoretically \textit{65536} different customers \cite{8716482}.

@@

    In the example Figure \ref{fig:a_plus_p} you can see that all internal ports are mapped to port values between \textit{4000} and \textit{4999}. This is because the ISP assigned this port range to that particular customer. All internal devices combined can open \textit{1000} ports. The biggest advantage of \textit{A+P} is also the biggest downside. On the one hand, it is possible to supply more customers with public, globally routable IPv4 address, but on the other hand all these customers have a tiny available port range. Assuming the extreme case of \textit{65536} different customers, each customer would only get one single port. Even when ports are dynamically assigned, the user experience at peak internet usage time would be unacceptable. Ports are important to distingush between different processes on a host. Each process that communicates with the network opens at least one socket and gets a port from the operating system. These ports are unsed to correctly deliver TCP or UDP packets. For example, a mail-programm opens a port to fetch the newest mails and a webbrowser opens ports to download different websites simultaniously. That is why each host needs multiple ports. Having multiple hosts in the network, which consume multiple ports each, can extend the provided port range. In case a customer is not supplied with sufficient ports by the ISP, the end-devices do not work as intended. 

@@

    \textit{NAT444} is also known as CGN. CGN means \textit{Carrier Grade Network Address Translation}. With NAT44 one public, globally routable IPv4 address is translated to multiple private IPv4 addresses at ISP site. Each customer gets one private IPv4 address. Similar to NAT44, the customer than translates that private IPv4 into multiple private IPv4 address for all home devices. That means there is a double NAT. IP addresses get translated twice \cite{10.1145/2987443.2987474}. This leads to even more problems than NAT44 or A+P \cite{8716482}.

@@

    In example Figure \ref{fig:nat_444} two end-users are connected to the CGN. The CGN has the public, globally routable IPv4 \textit{138.76.30.10}. One customer gets assigned the private IPv4 \textit{10.0.0.273} and the other one the private IPv4 \textit{10.0.0.274}. Both operate a typical \textit{192.168.0.0/16} Network. As you can see, both customers have a personal computer with the private IPv4 \textit{192.168.0.10}. Yet the system works. After the end-user NAT translation, these personal computers have the IPv4 \textit{10.0.0.273} and \textit{10.0.0.274} with the corresponding ports. Then the CGN translates these private IPv4 addresses into the only public, globally routable IPv4 \textit{138.76.30.10}. The CGN translates IP and Port and maps them to the corresponding customer. In a real world application, the CGN would control multiple public, gloablly routable IPv4 addresses and split them between all customers. Example Figure \ref{fig:nat_444} only shows a cutout of the whole CGN\cite{Hughes2022_C11}.

@@

    NAT44, A+P and NAT444 cause similar problems and have a huge impact on the end-user.\\\\The first problem is regarding location services. Typically it is possible to determine a rough geo-location regarding one single public IPv4. Especially with NAT444 one IPv4 is shared by multiple users. Thus it is not really possible to accurately guess the location. Some people might argue that this is a advantage too, because it increases privacy\cite{Hughes2022_C04}.\\\\A serious problem is regarding spam. IPv4 addresses which are distributing spam, typically can be blacklisted. With a shared IPv4 address between multiple end-users this can lead to blocking of innocent end-users who just were unlucky to share a IP with an adversary. In case a server only supports IPv4 and a end-user is behind a NAT444, the end-user has a worse experience if the server is using IP addresses for repuation. If a mallicious user misbehaves, an innocent user using the same public IPv4 can get blocked or rate-limited \cite{Hughes2022_C04}.\\\\Another problem is regarding peer-to-peer applications. To establish and maintain a peer-to-peer network it is necessary to ping a peer directly. This happens with a public IPv4 and a well known port. With NAT that is generally not possible. All devices share one IPv4 with all ports\cite{Hughes2022_C04}.\\\\Similar to peer-to-peer it is not possible to host own servers behind a NAT. Web servers for example establish \textit{https} connection over the well-known port \textit{443}. With NAT44 it is theoretically possible to forward a port to one device. Thus it is possible to operate one server of a kind per port. For example one web server and one ftp server. With A+P or NAT444 this option is completely gone and it is not possible to host servers. That impacts end-users massively. The main idea of the internet is a decentralised network. Everyone should be able to set up servers and start communications without a central instance. NAT disabled this main idea of the internet\cite{Hughes2022_C04}.\\\\Modern web applications nowadays usually require multiple ports to operate. With NAT these ports are limited and could run short. This is a serious problem, because as it is not possible to create more IPv4 addresses, it is not possible to create more Ports\cite{Hughes2022_C04}. \textit{A+P} makes this problem even worse, because the total available ports are even lower. As soon as all available ports are used, new processes cannot esatblish a internet connection\\\\ \textit{NAT444} could result in much fragmentation. Packets must travel from the end-device to the home router and from the home-router to the CGN. That are two fixed routes. Assuming that the first route can transport bigger packets, many packets need to be fragmented at the home router. This causes network overhead and performance degradation.\\\\Apart from the obvious disadvantages, some protocols like DNS are less secure with NAT. Because of the \textit{DNS poisoning attack}, a client starts new DNS requests with a random port. Assuming that an attacker cannot guess the port, the DNS server answers to that port. If an attacker can guess the port and answer to it first, the attacker can infiltrate wrong DNS entries. This is a serious security risk. With the use of NAT44 all devices share the available ports. The problem gets even worse with A+P or NAT444. If an attacker knows the assigned port range, the DNS cache poisoning attack is effective. IPSec needed to be upgraded to work with NAT44 and currently does not work with NAT444\cite{Hughes2022_C04}.\\\\A general problem with NAT is, that it is a single point of failure. If a personal NAT router or a CGN router is not working anymore, all clients cannot access the internet. A NAT router is a lucrative attack target\cite{Hughes2022_C04}.\\\\There are transport layer protocols that do not use ports at all. This is the case when multiplexing of the protcol is not needed. An example is the \textit{Interior gateway protocol} which is used to exchange routing information. Such protocols do not work with \textit{NAT}, because one IP address which should belong to one node is used by multiple nodes. This breaks the protocol.\\\\In a nutshell, NAT has problems. It was meant as a quick fix until the next generation of IP is finalized. Unfortunately, NAT outlived it supposed lifetime by years. This is why modern solutions are required to get away from NAT and return to the original open and accessable internet.

@@

    It became clear that IPv4 is not sufficient anymore. This is why IPv6 was standardized. IPv6 addresses are \textit{128} bit wide. With IPv6 it is possible to assign $2^{128}=3.4*10^{38}$ addresses. This number of addresses will suffice in the foreseable future. IPv6 solves the IPv4 exhaustion problem. Unfortunately IPv6 is not backwards compatible to IPv4. The first solution was to implement IPv6 alongside IPv4. This gives older applications the possibility to continue using IPv4 until they and the servers are upgraded. Assigning both an IPv4 address and IPv6 addresses is called \textit{dualstack}. The two similar technologies \textit{dualstack} and \textit{dualstack lite} are introduced now. They are meant to be transition technologies from the world of IPv4 only to IPv6 only.

@@

    With \textit{dualstack} the end-user gets a public, globally routable IPv4 address and a block IPv6 addresses from the ISP. With the IPv4 address, the end-user performs NAT44. This was discussed in the previous chapter. In addition to that private IPv4 address, all devices also get assigned a global, publicly routable IPv6 address \cite{8716482}. A personal computer now has two implemented IP layers as in Figure \ref{fig:dualstack}. Old application that only support IPv4 work seamlessly without any modification required.

@@

    With IPv6 an ISP typically get a \textit{/32} network containing $2^{128-32}=2^{96}$ IPv6 addresses. The end-user then gets a \textit{/64} network. IPv6 addresses are noted as 8 two byte blocks divides by a colon. The example in Figure \ref{fig:dualstack_network} shows a basic end-user \textit{dualstack} network containing one router and two end-devices. The ISP got the \textit{2001:0db8:0000:0000:0000:0000:0000:0000/32} block assigned. The router implements NAT44 and got the block \textit{2001:0db8:0001:0002:0000:0000:0000:0000/64} assigned by the ISP. The router can now distribute the last 4 two byte blocks to the devices. In this example the two personal computers got the public, globally routable IPv6 \textit{2001:0db8:0001:0002:0000:0000:0000:0010} and \textit{2001:0db8:0001:0002:0000:0000:0000:0012}. In addition to that the router assigns the private IPv4 addresses \textit{192.168.0.10} and \textit{192.168.0.12} to the personal computers. An application on one personal computer can now use the private IPv4 address with all the disadvantages discussed earlier or use the IPv6 address.

@@

        \caption{Example of a Dualstack network}

@@

    \textit{Dualstack Lite} also called \textit{DS Lite} is similar to \textit{dualstack} from a end-device point of view. An end-device gets a public, globally routable IPv6 address and a private IPv4 address. The end-device needs an \textit{dualstack} implementation as shown in Figure \ref{fig:dualstack}. In case the device wants to communicate with an IPv4 legacy server, the device has to use the IPv4 stack and the private IPv4. In all other cases the public, globally routable IPv6 is preferred. There is no translation between IPv4 and IPv6 or the other way around. All in all \textit{dualstack} and \textit{dualstack lite} does not differ for the end-device.\\\\Compared to \textit{dualstack}, the network traffic between the router of the customer and the ISP is IPv6 only. IPv4 packets are tunneled in IPv6. The IPv4 packet is part of the data field of the IPv6 packet. DNS is done over IPv6 only. The \textit{B4} element will perform DNS for all clients in the network. Thus DNS is never routed to \textit{AFTR}.\\\\Figure \ref{fig:dualstack_lite_network} is an example of a basic \textit{dualstack lite} setup. The customers router has a \textit{B4} element. This \textit{B4} element directly connects to the ISPs \textit{AFTR}. The \textit{B4} and \textit{AFTR} create an IPv6 tunnel for IPv4 packets. NAT44 is only applied at ISP level. Though, each customer hosts his own DHCPv4 Server to distribute IPv4 addresses in the home network. The IPv4 packets are packed into the IPv6 data field and directly send over IPv6 to the \textit{AFTR}. The \textit{AFTR} receives the IPv6 packets and extracts the IPv4 packets. Now the AFTR does NAT44. In addition to the original private IPv4 address and the port, the NAT table also saves the assigned IPv6 block for the customer. Because of that, the end-user can theoretically use an arbitrary private IPv4 block\cite{rfc6333}. Figure \ref{fig:dualstack_lite_network} shows that \textit{customer 1} and \textit{customer 2} both use the private IPv4 block \textit{192.168.0.0/16}. There are even two personal computer with both the private IPv4 \textit{192.168.0.12} in combination with the port \textit{3872}. This is no problem, because the NAT table also stores the customer assigned IPv6 block. \textit{Dualstack Lite} allows nearly endless scaling. It is not necessary to assign one public IPv4 per costumer. Over the time, the workload for the NAT table will decline because more and more clients will use the superior public, globally routable IPv6 address. There are only two cases, when IPv4 is needed. Either a legacy IPv4 application or device on the client side, or a device that wants to reach a server that is only recable over IPv4.\\\\\textit{Dualstack lite} also works with IPv4 only and IPv6 only devices. These only get assigned the corresponding IP address and work as intended. \textit{Dualstack lite} only uses NAT44 similar to \textit{dualstack}. That is why both are superior to \textit{NAT444}, while providing a usable solution even with IPv4 exhaustion\cite{Hughes2022_C08}.

@@

        \caption{Example of a dualstack Lite network}

@@

    Unfortunately \textit{dualstack} still requires the assignment of one or multiple IPv4 addresses per user. ISPs, especially mobile internet providers, run out of IPv4 addresses and want to transition to the new technology IPv6. This is mainly to reduce complexity and cut costs. That is why ISPs start to roll out IPv6 only connections. The end-user usually gets one \textit{/64} IPv6 block.\\\\If the end-user only accesses IPv6 content no problems appear. In case a end-device wants to access an IPv4 only server, network-based translation is required. Another problem occurs regarding IPv4 legacy applications that run on IPv6 only end-devices. These IPv4 requests must be translated too. This happens on the end-device itself with host-based translation.

@@

    \textit{6rd Rapid Deployment} is an extension to \textit{6in4}. It allows ISPs to limit the 6in4 tunnel to their own network. Thus they can provide access to all nodes in their network and guarantee good service.

>>>>

    \begin{abstract}
        This paper presents an overview of different techniques to cope with IPv4 address exhaustion. The impact on the end-users is on focus. First, there are some early NAT-based solutions introduced. Later dualstack is discussed. In the end, IPv6-only solutions deployed by ISPs today are presented. The end-user, especially the technically experienced one, can experience benefits with IPv6. With IPv6, the internet gets back to its roots. All devices can talk to each other, and central servers are not mandatory anymore to establish communication.
    \end{abstract}

@@

    In the late 90s, it became clear that the IPv4 address space was not sufficient to meet future demand. The popularity of the internet, especially in the end-user market, was heavily underestimated at the creation of IPv4. There are approximately 4 billion IPv4 addresses. Today there are over 8 billion humans on earth. That means that two humans must share a single IPv4 address, not considering servers. Today, a typical family has multiple devices per person, including TVs, gaming consoles, smartphones, personal computers, and more. This led to the assignment of the last available IPv4 block in the early 2010s. With the rise of smart home technology and IoT devices, the demand for IP addresses per household will continue increasing. In this paper, I will present a nearly complete overview of the different ideas to slow down the inevitable IPv4 exhaustion. Expect the last solution none of these is a long-term one or even tries to solve the problem itself. ISPs tried to keep the cost as low as possible. This led to keeping IPv4 addresses as long as possible.\\\\First, ISPs started to assign only one public IP per customer. The customer had to use NAT44 in order to provide all his end devices with an internet connection. The end-user was the one getting a worse experience. Over time, assigning only one IPv4 per customer was too much for the ISPs. Especially with the rise of mobile internet and an even higher demand for addresses, IPv4 was not sufficient anymore. ISPs tried to provide more IP addresses by assigning only private IPv4 addresses to customers, leading to NAT444. NAT itself has problems, which will be discussed in this paper. NAT444 makes the problems even worse.\\\\A few years later, dualstack was provided by ISPs. Dualstack introduced the next generation of IP addresses, called IPv6 addresses. Compared to IPv4 addresses, which are 32-bit wide, IPv6 addresses are 128-bit wide. IPv6 completely solves the problem of address exhaustion. Unfortunately, IPv6 is not backward compatible with IPv4. Thus all infrastructure needs to be renewed. The internet is fully decentralized. It is not possible to perform such a transition on one particular day. Dualstack implements both IPv4 and IPv6. Devices and servers supporting IPv6 can communicate over IPv6. In the case that one device is not yet capable of IPv6, IPv4 acts as a fallback. This is a perfect transition technology, but it is no solution for the IPv4 exhaustion problem. Each end-user still needs a public or private IPv4. The problem is not solved.\\\\Lately, ISPs have started to provide IPv6 only. This is also common in mobile networks. The devices natively speak to servers over IPv6. For legacy IPv4 servers, there are different translation mechanisms. This paper will take a closer look into NAT64 combined with DNS64. Moreover, I will present 464XLAT. Both are similar to a NAT44 but translate between IPv6 and IPv4. The exciting thing is that these translations happen at the ISP, so the end-user only uses IPv6. This has advantages for the end-user. Though, some legacy applications that depend on IPv4 cannot be used. For example, GitHub is not reachable over Ipv6. The DNS does not return a \textit{AAAA} record.\\\\At the end of this paper, tunneling IPv6 in IPv4 is presented. This happens only at the ISP level and indirectly influences the end-user too. Mainly in better path availability, thus seamless internet service. There are various technologies. 6in4 and 6rd will be introduced shortly.\\\\In the last few years, the adoption of IPv6 has increased rapidly. Above 40 percent of all google users access the site over Ipv6 \cite{GoogleIPv6}. These statistics show the trend of ISPs worldwide deploying IPv6 only or dualstack. Nearly all mobile clients use IPv6 today and can interact with the internet without notable disadvantages. Cisco predicts that the transition to IPv6 will be finished in 2028 \cite{Hughes2022_C05}. This means that nearly all end-users and servers worldwide communicate over IPv6. That has noticeable advantages for the end-user regarding security because firewalls can be managed easier. In addition, it opens up entirely new options for innovative peer-to-peer technologies. Hosting a Server on a smartphone might sound strange today, but it could be a usual thing ten years from now. The end-user is the one benefiting the most from unique, worldwide routable IP addresses.\\\\Now I am going to turn back the time a bit, and we are focusing on the problem emerging in the late 90s. It becomes clear that the IPv4 address space is not big enough. ISPs are searching for a solution. The main goal is to assign the same public, globally routable IPv4 address to multiple devices. This is not meant for an extended period but much more as a quick fix until the new standard IPv6 is developed.

@@

    This section presents three different solutions to extend the lifetime of IPv4: NAT44, A+P, and NAT444. All these solutions aim to share one single IPv4 address between multiple end devices. It is essential to understand these technologies in order to estimate the impact of IPv6 deployment by ISPs correctly. Evaluating modern solutions alone could lead to the impression that these solutions are better or worse than they are. Thus this paper starts with NAT44.

@@

    Everybody can use the IPv4 addresses in these blocks. Private IPv4 addresses will never be routed on the internet. Most end-users use the third option for their home network, but only with a /24 network. The third block remains 0 all the time. Thus they can use 254 different devices, which is sufficient for most users. All these 254 devices share one public IPv4. All of them share the $2^{16} = 65536$ ports of the public IPv4 too. From the outside, one cannot decide whether only one device or multiple devices are behind the NAT.

@@

    In the example, Figure \ref{fig:nat_44} is a NAT-Translation table. As one can see, the PC with IP \textit{192.168.0.10} cannot use all ports because the PC with the IP \textit{192.168.0.12} uses the Port \textit{7035}. All ports are shared. This can lead to problems, as we will discuss later. NAT44 was the first solution to the IPv4 exhaustion problem. It is an easy fix because the ISP has nothing to change, except now only delivering one single IP per customer. The end-user needs a NAT-compatible router.

@@

    \textit{A+P} means \textit{Address Plus Port}. Over time NAT44 was not a sufficient solution anymore. The demand for IPv4 rose, and ISP were not able to provide one single public, globally routable IPv4 per customer. The idea is still similar to NAT44, but with \textit{A+P} each customer gets one IPv4 and a designated port range. With that system, it is possible to share one IP to theoretically \textit{65536} different customers \cite{8716482}.

@@

    In the example Figure \ref{fig:a_plus_p} you can see that all internal ports are mapped to port values between \textit{4000} and \textit{4999}. This is because the ISP assigned this port range to that particular customer. All internal devices combined can open \textit{1000} ports. The most significant advantage of \textit{A+P} is also the most significant downside. On the one hand, it is possible to supply more customers with public, globally routable IPv4 addresses, but on the other hand, all these customers have a tiny available port range. Assuming the extreme case of \textit{65536} different customers, each customer would only get one single port. Even when ports are dynamically assigned, the user experience at peak internet usage time would be unacceptable. Ports are essential to distinguish between different processes on a host. Each process that communicates with the network opens at least one socket and gets a port from the operating system. These ports are used to deliver TCP or UDP packets correctly. For example, a mail program opens a port to fetch the newest mail, and a web browser opens ports to download different websites simultaneously. That is why each host needs multiple ports. Having multiple hosts in the network, which consume multiple ports each, can extend the provided port range. In case a customer is not supplied with sufficient ports by the ISP, the end devices do not work as intended. 

@@

    \textit{NAT444} is also known as CGN. CGN means \textit{Carrier Grade Network Address Translation}. With NAT44, one public, globally routable IPv4 address is translated to multiple private IPv4 addresses at the ISP site. Each customer gets one private IPv4 address. Similar to NAT44, the customer then translates that private IPv4 into multiple private IPv4 address for all home devices. That means there is a double NAT. IP addresses get translated twice \cite{10.1145/2987443.2987474}. This leads to even more problems than NAT44 or A+P \cite{8716482}.

@@

    In Figure \ref{fig:nat_444} two end-users are connected to the CGN. The CGN has the public, globally routable IPv4 \textit{138.76.30.10}. One customer gets assigned the private IPv4 \textit{10.0.0.273} and the other the private IPv4 \textit{10.0.0.274}. Both operate a typical \textit{192.168.0.0/16} network. As one can see, both customers have a personal computer with the private IPv4 \textit{192.168.0.10}. Nevertheless, the system works. After the end-user NAT translation, these personal computers have the IPv4 \textit{10.0.0.273} and \textit{10.0.0.274} with the corresponding ports. Then the CGN translates these private IPv4 addresses into the only public, globally routable IPv4 \textit{138.76.30.10}. Finally, the CGN translates IP and Port and maps them to the corresponding customer. In a real-world application, the CGN would control multiple public, globally routable IPv4 addresses and split them between all customers. Example Figure \ref{fig:nat_444} only shows a cutout of the whole CGN\cite{Hughes2022_C11}.

@@

    NAT44, A+P, and NAT444 cause similar problems and have a significant impact on the end-user.\\\\The first problem is regarding location services. Typically it is possible to determine a rough geo-location regarding one single public IPv4. Especially with NAT444, one IPv4 is shared by multiple users. Thus it is not possible to accurately guess the location. Some people might argue that this is an advantage too because it increases privacy\cite{Hughes2022_C04}.\\\\A serious problem is regarding spam. IPv4 addresses which are distributing spam get typically denylisted. With one shared IPv4 address between multiple end-users, this can lead to the blocking of innocent end-users who just were unlucky to share an IP with an adversary. In case a server only supports IPv4, and an end-user is behind a NAT444, the end-user has a worse experience if the server is using IP addresses for reputation. If a malicious user misbehaves, an innocent user using the same public IPv4 can get blocked, or rate-limited \cite{Hughes2022_C04}.\\\\Another problem is regarding peer-to-peer applications. To establish and maintain peer-to-peer networks, it is necessary to ping peers directly. This happens with a public IPv4 and a well-known port. With NAT, that is generally not possible. All devices share one IPv4 with all ports\cite{Hughes2022_C04}.\\\\Similar to peer-to-peer, it is not possible to host your servers behind a NAT. Web servers, for example, establish \textit{https} connection over the well-known port \textit{443}. With NAT44, it is theoretically possible to forward a port to one device. Thus it is possible to operate one server of a kind per port, for example, one web server and one FTP server. With A+P or NAT444, this option is completely gone, and it is not possible to host servers. That impacts end-users massively. The main idea of the internet is a decentralized network. Everyone should be able to set up servers and start communications without a central instance. NAT disabled this main idea of the internet\cite{Hughes2022_C04}.\\\\Modern web applications nowadays usually require multiple ports to operate. With NAT, these ports are limited and could run short. This is a serious problem because as it is not possible to create more IPv4 addresses, it is not possible to create more Ports\cite{Hughes2022_C04}. \textit{A+P} makes this problem even worse because the total available ports are even lower. As soon as all available ports are used, new processes cannot establish an internet connection\\\\ \textit{NAT444} could result in much fragmentation. Packets must travel from the end device to the home router and from the home router to the CGN. These are two fixed routes. Assuming that the first route can transport bigger packets, many packets need to be fragmented at the home router. This causes network overhead and performance degradation.\\\\Apart from the apparent disadvantages, some protocols like DNS are less secure with NAT. Because of the \textit{DNS poisoning attack}, a client starts new DNS requests with a random port. Assuming that an attacker cannot guess the port, the DNS server answers that port. If an attacker can guess the port and answer to it first, the attacker can infiltrate wrong DNS entries. This is a severe security risk. With the use of NAT44, all devices share the available ports. The problem gets even worse with A+P or NAT444. If an attacker knows the assigned port range, the DNS cache poisoning attack is effective. IPSec needed to be upgraded to work with NAT44 and currently does not work with NAT444\cite{Hughes2022_C04}.\\\\A general problem with NAT is that it is a single point of failure. If a personal NAT router or a CGN router is not working anymore, all clients cannot access the internet. A NAT router is a lucrative attack target\cite{Hughes2022_C04}.\\\\There are transport layer protocols that do not use ports at all. This is the case when multiplexing of the protocol is not needed. An example is the \textit{Interior gateway protocol} which is used to exchange routing information. Such protocols do not work with \textit{NAT}, because multiple nodes use one IP address that should belong to one node. This breaks the protocol.\\\\In a nutshell, NAT has problems. It was meant as a quick fix until the next generation of IP is finalized. Unfortunately, NAT outlived its supposed lifetime by years. This is why modern solutions are required to get away from NAT and return to the original open and accessible internet.

@@

    It became clear that IPv4 is not sufficient anymore. This is why IPv6 was standardized. IPv6 addresses are \textit{128} bit wide. With IPv6 it is possible to assign $2^{128}=3.4*10^{38}$ addresses. This number of addresses will suffice in the foreseeable future. IPv6 solves the IPv4 exhaustion problem. Unfortunately, IPv6 is not backward compatible with IPv4. The first solution was to implement IPv6 alongside IPv4. This gives older applications the possibility to continue using IPv4 until they and the servers are upgraded. Assigning both an IPv4 address and an IPv6 address is called \textit{dualstack}. The two similar technologies \textit{dualstack} and \textit{dualstack lite} are introduced now. They are meant to be transition technologies from the world of IPv4 only to IPv6 only.

@@

    With \textit{dualstack} the end-user gets a public, globally routable IPv4 address and a block IPv6 address from the ISP. With the IPv4 address, the end-user performs NAT44. This was discussed in the previous chapter. In addition to that private IPv4 address, all devices also get assigned a global, publicly routable IPv6 address \cite{8716482}. A personal computer now has two implemented IP layers as in Figure \ref{fig:dualstack}. An old application that only supports IPv4 works seamlessly without any modification required.

@@

    With IPv6 an ISP typically get a \textit{/32} network containing $2^{128-32}=2^{96}$ IPv6 addresses. The end-user then gets a \textit{/64} network. IPv6 addresses are noted as eight two-byte blocks divided by a colon. The example in Figure \ref{fig:dualstack_network} shows a basic end-user \textit{dualstack} network containing one router and two end devices. The ISP got the \textit{2001:0db8:0000:0000:0000:0000:0000:0000/32} block assigned. The router implemented NAT44 and got the block \textit{2001:0db8:0001:0002:0000:0000:0000:0000/64} assigned by the ISP. The router can now distribute the last four two-byte blocks to the devices. In this example, the two personal computers got the public, globally routable IPv6 \textit{2001:0db8:0001:0002:0000:0000:0000:0010} and \textit{2001:0db8:0001:0002:0000:0000:0000:0012}. In addition to that, the router assigns the private IPv4 addresses \textit{192.168.0.10} and \textit{192.168.0.12} to the personal computers. An application on one personal computer can now use the private IPv4 address with all the disadvantages discussed earlier or use the IPv6 address.

@@

        \caption{example of a Dualstack network}

@@

    \textit{Dualstack Lite} also called \textit{DS Lite} is similar to \textit{dualstack} from a end-device point of view. An end device gets a public, globally routable IPv6 address, and a private IPv4 address. The end-device needs an \textit{dualstack} implementation as shown in Figure \ref{fig:dualstack}. In case the device wants to communicate with an IPv4 legacy server, the device has to use the IPv4 stack and the private IPv4. In all other cases, the public, globally routable IPv6 is preferred. There is no translation between IPv4 and IPv6 or the other way around. All in all, \textit{dualstack} and \textit{dualstack lite} do not differ for the end-device.\\\\Compared to \textit{dualstack}, the network traffic between the router of the customer and the ISP is IPv6 only. IPv4 packets are tunneled in IPv6. The IPv4 packet is part of the data field of the IPv6 packet. DNS is done over IPv6 only. The \textit{B4} element will perform DNS for all clients in the network. Thus DNS is never routed to \textit{AFTR}.\\\\Figure \ref{fig:dualstack_lite_network} is an example of a basic \textit{dualstack lite} setup. The customer's router has a \textit{B4} element. This \textit{B4} element directly connects to the ISPs \textit{AFTR}. The \textit{B4} and \textit{AFTR} create an IPv6 tunnel for IPv4 packets. NAT44 is only applied at the ISP level. Though each customer hosts his DHCPv4 server to distribute IPv4 addresses in the home network. The IPv4 packets are packed into the IPv6 data field and directly sent over IPv6 to the \textit{AFTR}. The \textit{AFTR} receives the IPv6 packets and extracts the IPv4 packets. Now the AFTR does NAT44. In addition to the original private IPv4 address and the port, the NAT table also saves the assigned IPv6 block for the customer. Because of that, the end-user can theoretically use an arbitrary private IPv4 block\cite{rfc6333}. Figure \ref{fig:dualstack_lite_network} shows that \textit{customer 1} and \textit{customer 2} both use the private IPv4 block \textit{192.168.0.0/16}. There are even two personal computers that are assigned both with the private IPv4 \textit{192.168.0.12} in combination with the port \textit{3872}. This is not a problem because the NAT table also stores the customer-assigned IPv6 block. \textit{Dualstack Lite} allows nearly endless scaling. It is not necessary to assign one public IPv4 per customer. Over time, the workload for the NAT table will decline because more and more clients will use the superior public, globally routable IPv6 address. There are only two cases when IPv4 is needed. Either a legacy IPv4 application or device on the client side or a device that wants to reach a server that is only reachable over IPv4.\\\\\textit{Dualstack lite} also works with IPv4 only and IPv6 only devices. These only get assigned the corresponding IP address and work as intended. \textit{Dualstack lite} only uses NAT44 similar to \textit{dualstack}. That is why both are superior to \textit{NAT444}, while providing a usable solution even with IPv4 exhaustion\cite{Hughes2022_C08}.

@@

        \caption{example of a dualstack Lite network}

@@

    Unfortunately, \textit{dualstack} still requires the assignment of one or multiple IPv4 addresses per user. ISPs, especially mobile internet providers, run out of IPv4 addresses and want to transition to the new technology, IPv6. This is mainly to reduce complexity and cut costs. That is why ISPs have started to roll out IPv6-only connections. The end-user usually gets one \textit{/64} IPv6 block.\\\\If the end-user only accesses IPv6 content, no problems appear. However, if an end device wants to access an IPv4-only server, network-based translation is required. Another problem occurs regarding IPv4 legacy applications that run on IPv6-only end devices. These IPv4 requests must be translated too. This happens on the end device itself with host-based translation.

@@

    \textit{6rd Rapid Deployment} is an extension to \textit{6in4}. It allows ISPs to limit the 6in4 tunnel to their network. Thus they can provide access to all nodes in their network and guarantee good service.
